---
title: "HW1: Mouse Body Weight. t-Test, ANOVA and dot-plot examples"
output: html_notebook
---

# Load R libraries
```{r setup}
require(data.table)
require(ggplot2)
require(DT)
require(multcomp)
```

# Experimental setup
Treatments:  
**Naive** - healthy mice  
** Negative Control** - mice treated with lipopolysaccharide (LPS) to induce gut inflammation that leads to weight loss. These mice do not receive any treatment.  
**Pontikilin, Choirídin and Tipotium** - treatments to prevent the effect of LPS. The treatments are administered to mice previously treated with LPS.  
  
Diet:  
**Grass** - normal diet  
**Grass and Carrot** - normal diet plus carrots  
  
Mice:
All 60 mice are female, same age and genotype. If they were not treated, they would all be expected to be similar to the Naive group, weighting around 20g at the time of measurements.  

# Simulate data
```{r data_sim}
set.seed(1110198)
dt1 <- data.table(ID = 1:60,
                  Treatment = rep(rep(c("Naive",
                                        "Negative Control",
                                        "Pontikilin",
                                        "Choirídin",
                                        "Tipotium"),
                                      each = 6),
                                  2),
                  Diet = rep(c("Grass",
                               "Grass and Carrot"),
                             30))
dt1$ID <- factor(dt1$ID,
                 levels = unique(dt1$ID))

dt1$Treatment <- factor(dt1$Treatment,
                        levels = unique(dt1$Treatment))

dt1$Diet <- factor(dt1$Diet,
                   levels = unique(dt1$Diet))


dt1[, Weight := round(20 -
                        5* (Treatment == "Negative Control") -
                        1*(Treatment == "Pontikilin")  - 
                        4*(Treatment == "Pontikilin" & Diet == "Grass and Carrot") -
                        3*(Treatment == "Choirídin") - 
                        5*(Treatment == "Tipotium") -
                        # 2*(Diet == "Grass") +
                        rnorm(n = 60,
                              mean = 0,
                              sd = 1),
                      2)]

save(dt1,
     file = "data/dt1.RData")
write.csv(dt1,
          file = "data/dt1.CSV",
          row.names = FALSE)

datatable(dt1,
          rownames = FALSE)
```

# Read saved RData (HW1)
```{r data}
load("data/dt1.RData")
dt1
```

# Plot data
```{r dot_plot}
p1 <- ggplot(dt1,
             aes(x = Treatment,
                 y = Weight,
                 group = ID,
                 fill = Diet)) + 
  geom_point(shape = 21,
             size = 3,
             position = position_dodge(0.3)) +
  scale_y_continuous("Weight (gm)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1))

# Save the plot as a TIFF file
tiff(filename = "tmp/dotplot.tiff",
     height = 5,
     width = 6,
     units = 'in',
     res = 300,
     compression = "lzw+p")
print(p1)
graphics.off()

print(p1)
```
# t-Test: controls only
```{r}
t.test(Weight ~ Treatment,
       data = dt1[Treatment %in% c("Naive",
                                   "Negative Control"), ],
       var.equal = TRUE)
```

# Wilcoxon test
```{r}
wilcox.test(Weight ~ Treatment,
       data = dt1[Treatment %in% c("Naive",
                                   "Negative Control"), ])
```

# Boxplot
```{r}
p2 <- ggplot(dt1,
             aes(x = Treatment,
                 y = Weight,
                 fill = Treatment)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        legend.position = "none")

# Save the plot as a TIFF file
tiff(filename = "tmp/boxplot.tiff",
     height = 5,
     width = 5,
     units = 'in',
     res = 300,
     compression = "lzw+p")
print(p2)
graphics.off()
p2
```

# ANOVA
```{r}
m1 <- aov(Weight ~ Treatment,
          data = dt1)
summary(m1)
```
# Pairwise comparison
```{r}
# p-Values adjusted for multiplicity
m1_mult_comp_adj <- glht(m1,
                         linfct = mcp("Treatment" = "Tukey"))
m1_mult_comp_adj
summary(m1_mult_comp_adj)
```

# Examples of t-Test and ANOVA
You need to adopt it to analyze **your** data. Be mindful of different groups and data structure in general. Read up the documentation and understand what the functions do and how the tests are done.  
  
## t-Test example
```{r t_test}
?t.test

set.seed(1110198)
dt2 <- data.table(Group = rep(c("A", "B"), 
                              each = 30),
                  Readout = c(rnorm(n = 30,
                                    mean = 10,
                                    sd = 5),
                              rnorm(n = 30,
                                    mean = 15,
                                    sd = 5)))
head(dt2)
                  
p1 <- ggplot(dt2,
             aes(x = Readout,
                 group = Group,
                 fill = Group)) +
  geom_histogram(color = "black",
                 bins = 6,
                 alpha = 0.7)
p1

p2 <- ggplot(dt2,
             aes(x = Group,
                 y = Readout,
                 group = Group,
                 fill = Group)) +
  geom_boxplot()
p2

t.test(Readout ~ Group,
       data = dt2)
```

## Analysis Of VAriance (ANOVA)
```{r anova}
set.seed(1110198)
dt3 <- data.table(Group = factor(rep(c("A", "B", "C"), 
                                     each = 30)),
                  Readout = c(rnorm(n = 30,
                                    mean = 10,
                                    sd = 5),
                              rnorm(n = 30,
                                    mean = 15,
                                    sd = 5),
                              rnorm(n = 30,
                                    mean = 8,
                                    sd = 5)))
head(dt3)

p3 <- ggplot(dt3,
             aes(x = Readout,
                 group = Group,
                 fill = Group)) +
  geom_histogram(color = "black",
                 bins = 6,
                 alpha = 0.7)
p3

p4 <- ggplot(dt3,
             aes(x = Group,
                 y = Readout,
                 group = Group,
                 fill = Group)) +
  geom_boxplot()
p4

?aov
m1 <- aov(Readout ~ Group,
            data = dt3)
summary(m1)

?multcomp::glht
# p-Values adjusted for multiplicity
m1_mult_comp_adj <- glht(m1,
                         linfct = mcp("Group" = "Tukey"))
m1_mult_comp_adj
summary(m1_mult_comp_adj)
```

# Basic stats
```{r}
dt1 <- data.table(Observation = 1:5,
                  X = c(3, 1, 12, 2, 7))

p1 <- ggplot(dt1,
       aes(x = Observation,
           y = X,
           fill = factor(Observation))) +
  geom_hline(color = "red",
             yintercept = mean(dt1$X),
             size = 1) +
  geom_hline(color = "blue",
             yintercept = median(dt1$X),
             linetype = "dashed") +
  geom_segment(aes(y = c(3, 1, 5, 2, 5),
                   yend = c(5, 5, 12, 5, 7),
                   x = Observation,
                   xend = Observation)) +
  geom_text(aes(x = 2, 
                y = 5.5,
                label = "mean",
                color = "blue"),
            size = 5) +
  geom_text(aes(x = 3, 
                y = 3.5,
                label = "median",
                color = "red"),
            size = 5) +
  geom_point(size = 6,
             shape = 21) +
  scale_y_continuous(breaks = 1:13) +
  theme_bw() +
  theme(legend.position = "none")

# Save the plot as a TIFF file
tiff(filename = "tmp/mean_med.tiff",
     height = 4,
     width = 4,
     units = 'in',
     res = 300,
     compression = "lzw+p")
print(p1)
graphics.off()

print(p1)
```

```{r sd_se}
set.seed(11111)
dt4 <- data.table(ID = 1:110,
                  Y = c(rnorm(n = 10, mean = 100, sd = 10),
                        rnorm(n = 100, mean = 100, sd = 10)),
                  X = c(rep("10 Samples", 10),
                        rep("100 Samples", 100)))

p4 <- ggplot(dt4,
             aes(x = X,
                 y = Y,
                 group = ID,
                 fill = X)) + 
  geom_hline(yintercept = 100,
             linetype = "dashed") +
  geom_point(shape = 21,
             size = 3,
             position = position_dodge(0.3)) +
  scale_x_discrete("") +
  scale_y_continuous("Value") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1),
        legend.position = "none")

# Save the plot as a TIFF file
tiff(filename = "tmp/sd_se.tiff",
     height = 4,
     width = 4,
     units = 'in',
     res = 300,
     compression = "lzw+p")
print(p4)
graphics.off()

print(p4)
```
# SD and SE
```{r}
dt4[ ,.(SD = sd(Y),
       SEM = sd(Y)/sqrt(.N)),
by = X]
```
